---
title: "Visual analytics of hotel bookings data"
author: "Julià Minguillón"
date: "April 2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

NOTE: this tutorial uses R + RStudio + some R packages to show the potential
of using data visualization for inspecting and analyzing a data set. We strongly
recommend you to explore the following links:

1) RStudio: https://posit.co/downloads/
2) ggplot2: https://ggplot2.tidyverse.org/
3) extensions: https://exts.ggplot2.tidyverse.org/gallery/
4) ggmosaic: this package has been removed from CRAN, it is necessary to install
and older version:

Download and install RTools from "https://cran.rstudio.com/bin/windows/Rtools/rtools45/rtools.html"

Download ggmosaic running install.packages( "ggmosaic", repos = c("https://haleyjeppson.r-universe.dev", "https://cloud.r-project.org")))

## Load packages

```{r packages}
library("ggmosaic")
library("ggplot2")
library("fitdistrplus")
library("MASS")
library("survival")
library("ggstatsplot")
library("tidyverse")
library("lubridate")
```

## Data loading and dimensions (N x M)

We read the dataset in CSV format, with 119,390 rows y 32 columns:

```{r load data}
x=read.csv("hotel_bookings.csv", stringsAsFactors = T)
dim(x)
```
## Data cleansing

First, we'll inspect the data using the summary() function included in R.  You 
can find an explanation of each variable in the article that describes this 
dataset in detail, although the variable names are pretty much self-explanatory:

```{r summary, echo=FALSE}
summary(x)
```

# Numerical variables

Some unexpected (outliers?) values for several variables can be observed.
For instance:

1) A maximum of 55 in 'adults'
2) A maximum of 10 in 'children' (including also missing values)
3) A maximum of 10 in 'babies'
4) Negative values in the average daily rate ('adr') or or very high

Let's visualize the histogram of the variable 'adults', with at least 55 
breaks in the histogram, using the function hist() in R:

```{r hist_adults}
hist(x$adults,breaks=55)
```

It can be observed that the histogram shows no bars around the value 55, 
given that this is a very large set and probably it's only one or a few cases. 
In these cases, to analyze the extreme values of a variable, the values of the 
variable in question can be represented graphically as follows, ordering and 
plotting the data (if they are numerical, as in this case):

```{r plot_adults}
plot(sort(x$adults))
grid()
```
The 'Index' represents the position of the element once it's sorted, but we're 
more interested in the Y axis, as we can see that some elements have values 
of 10 or higher. Since this is an integer variable with a limited set of 
possible values, we can use table() to visualize them:

```{r table_adults}
table(x$adults)
```
As you can see, there's one reservation for 10 adults, two for 20 adults, 
and so on, up to one for 55 adults! Without going into further detail, we'll 
remove all rows with reservations for 10 or more adults:

```{r clean_adults}
x=x[x$adults<10,]
```

EXERCISE: Repeat this process with variables 'children' and 'babies'. Try also
to change the threshold to less than 5 instead of 10. 

```{r exercise_clean_guests}
# Removing NAs for children before filtering to avoid losing data unexpectedly
x$children[is.na(x$children)] <- 0

# Apply threshold of 5 for children and babies
x <- x[x$children < 5, ]
x <- x[x$babies < 5, ]

# Verify results
table(x$children)
table(x$babies)
```

The histogram of the 'adr' variable (average daily rate) presents the same 
problem as the 'adults' variable, so we will simply create a graph with the 
ordered values again:

```{r plot_adr}
plot(sort(x$adr))
grid()
```
In this case, we observe that only one value is significantly higher than the 
rest. We consider it an outlier and eliminate it, as well as the negative 
values which have no a clear explanation, although we keep the 0 values:

```{r clean_adr}
x=x[x$adr>=0 & x$adr<1000,]
```

The histogram now provides us with some relevant information. We draw it using
the ggplot2 package, which offers many more options than hist():

```{r hist_adr}
ggplot(data=x, aes(x=adr)) + 
  geom_histogram(bins=55, colour="black", fill = "lightgray") +
  theme_light()
```
EXERCISE: improve the graph to make axis, title, etc. more adequate.

```{r exercise_improved_hist}
ggplot(data=x, aes(x=adr)) + 
  geom_histogram(bins=60, colour="white", fill = "#2c3e50") +
  labs(title = "Distribution of Average Daily Rate (ADR)",
       subtitle = "Values filtered below 1000€",
       x = "Average Daily Rate (EUR)",
       y = "Frequency",
       caption = "Source: Hotel Bookings Dataset") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))
```

We can see that there is a set of approximately 2,000 zero values, which could
be analyzed separately, for example. There are R packages that help us estimate
this distribution and the parameters that determine it visually, such as the 
fitdistrplus package, which provides the descdist() function (caution, slow!):

```{r descdist}
require(fitdistrplus)
descdist(x$adr,boot=1000)
```
As you can see, the real data (observations, a colored dot) and the simulated
data (in other color) approximate what a lognormal distribution might look like. 
However, to experiment with the cleanest possible data set, we will:

1) remove 0-day stays
2) remove 0-cost stays
3) remove stays with no guests
4) replace the NAs in the children variable with 0

```{r data_cleansing}
x[is.na(x$children),'children']=0
x=x[x$adr>0 & 
    (x$stays_in_week_nights+x$stays_in_weekend_nights)>0 & 
    (x$adults+x$children+x$babies)>0 & 
    !is.na(x$children),]
```

## Categorical variables

For categorical variables, the summary() function gives us a first idea of the
possible values each can take. For example, in the original set (before removing
outliers), there are 79,330 reservations at a city hotel (Lisbon) and 40,060 at
a resort (Algarve). We can ask ourselves whether the cost distribution is the
same for both groups, either by using the appropriate statistical test or simply
by comparing histograms, in this case using the ggplot2 package, which is much
more powerful for creating all kinds of graphs:

```{r hist_adr_tipo}
# require(ggplot2)
ggplot(data=x, aes(x=adr, fill=hotel)) + 
  geom_histogram(bins=50, colour="black") +
  theme_light()
```
It can be seen that the most common prices in Lisbon (city hotels) are slightly
to the right of the most common prices in the Algarve (resort hotels), although
the highest prices in Lisbon decrease more rapidly than in the Algarve. By using
a violin plot, we can see more detail, especially if we also show the typical
quartiles of a box plot:

```{r violin_adr_tipo}
ggplot(data=x, aes(x=hotel, y=adr, fill=hotel)) + 
  geom_violin() + geom_boxplot(width=.1, outliers = F) +
  coord_flip() + 
  theme_light()
```
There is an R package called ggstatsplot that has specific functions for each
type of graph, including appropriate statistical tests to determine if there
are differences between groups:

```{r ggstatsplot}
# require(ggstatsplot)
ggbetweenstats(data=x, x=hotel, y=adr)
```
Another interesting variable is the hotel guests' origin ('country'). The
problem is that this variable has many different values (178), so we should
focus on the countries with the most tourists, also showing whether they choose
a city hotel or a resort:

```{r country}
# require(tidyverse)
# countries with at least 100 bookings
xx = x %>% group_by(country) %>% mutate(pais=n()) %>% filter(pais>=100)
xx$country=factor(xx$country)
ggplot(data=xx, aes(x=reorder(country, -pais))) + 
  geom_bar(stat="count", aes(fill=hotel)) +
  theme_light() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```
Obviously, Portugal (PRT) ranks first, followed by neighboring countries such
as Great Britain, France, and Spain. Visitors from Great Britain and Ireland
are most likely to choose a resort, while those from France, Germany, and Italy
primarily visit Lisbon.

EXERCISE: Are there differences between residents of Portugal and the rest?

```{r exercise_portugal_rest}
# Create a new grouping variable
x$is_portuguese <- ifelse(x$country == "PRT", "Portugal", "International")

# Compare ADR between Portuguese and international guests
ggbetweenstats(data=x, x=is_portuguese, y=adr, 
               title = "ADR Comparison: Local vs International Guests")
```

Another interesting variable is 'is_canceled', which indicates whether a 
reservation was canceled or not (37.0% of the time). We can observe the 
relationship between two categorical variables using a mosaic chart:

```{r mosaic_hotel_is_canceled}
# require(ggmosaic)
x$is_canceled=as.factor(x$is_canceled)
ggplot(data=x) + 
  geom_mosaic(aes(x=product(is_canceled, hotel), fill=hotel)) +
  theme_light() 
```
It can be seen that the cancellation rate (denoted by 1 on the Y-axis) at a 
resort is lower than that of a hotel in Lisbon. On the X-axis, the relative
size of each column also corresponds to the proportion of each hotel type. 
It is important not to consider the Y-axis labels (0/1) as the actual numerical
cancellation rate, as this can be misleading.

EXERCISE: which other type of graph could be used to represent this data?

```{r exercise_alternative_categorical}
# A stacked bar chart (normalized to 100%) is a clear alternative
ggplot(x, aes(x = hotel, fill = is_canceled)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Cancellation Proportions by Hotel Type",
       y = "Percentage", fill = "Canceled (1=Yes)") +
  theme_minimal()
```

In the case of cancellation by country for the countries with more tourists:

```{r mosaic_country_is_canceled}
# at least 1000 bookings
xx = x %>% group_by(country) %>% mutate(pais=n()) %>% filter(pais>=1000)
xx$country=factor(xx$country)
ggplot(data=xx) + 
  geom_mosaic(aes(x=product(is_canceled, country), fill=country)) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```
It can be seen that the cancellation rate is much higher for local tourists 
(from Portugal, PRT), while it is much lower for the rest of the countries.
However, this graph is not easy to read; in this case, there is no order of 
either the countries or the percentage of cancellations.

EXERCISE: Improve the previous graph to make it more understandable and 
consider whether it is possible to visualize the relationships between three
or more categorical variables.

```{r exercise_improved_categorical}
# To make it understandable, we calculate cancellation rates and reorder
xx_stats <- x %>% 
  group_by(country) %>% 
  summarise(n = n(), 
            cancel_rate = mean(as.numeric(as.character(is_canceled)))) %>% 
  filter(n >= 1000)

ggplot(xx_stats, aes(x = reorder(country, -cancel_rate), y = cancel_rate, fill = cancel_rate)) +
  geom_col() +
  scale_fill_gradient(low = "steelblue", high = "red") +
  labs(title = "Cancellation Rate by Country (Top Countries)",
       x = "Country", y = "Cancellation Rate") +
  theme_light()

# Visualizing 3 variables (Country, Hotel, Cancellation)
xx_3var <- x %>% 
  group_by(country, hotel) %>% 
  mutate(n = n()) %>% 
  filter(n >= 500)

ggplot(xx_3var, aes(x = country, fill = is_canceled)) +
  geom_bar(position = "fill") +
  facet_wrap(~hotel) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Cancellations by Country and Hotel Type")
```

Finally, let's analyze the behavior of reservations relative to the arrival
date. First, using the R lubridate package (a marvel for manipulating date and
time data), we'll create a 'day' variable to determine the day of the week the
hotel was checked in and analyze how many reservations there were each day:

```{r dia}
# require(lubridate)
x$dia=as_date(paste0(x$arrival_date_year,'-',x$arrival_date_month,'-',x$arrival_date_day_of_month))
ggplot(data=x,aes(x=dia,group=arrival_date_year,color=as.factor(arrival_date_year))) + 
  geom_bar() + scale_color_manual(values=c("2015"="red","2016"="green","2017"="blue")) + 
  theme_light() + 
  theme(legend.position='none') 
```
EXERCISE: Improve and split the above graph by hotel type or country of origin.

```{r exercise_date_split}
# Splitting by hotel type using facet_wrap for better clarity
ggplot(data=x, aes(x=dia, color=hotel)) + 
  geom_line(stat="count") + 
  facet_wrap(~hotel, scales = "free_y", ncol = 1) +
  labs(title = "Daily Arrivals by Hotel Type", x = "Date", y = "Number of Bookings") +
  theme_light()
```

As described in the article, the data covers the period from July 1, 2015, to
August 31, 2017. Some peaks can be observed that might be interesting to explain
(what happened those days, i.e. 2015-12-05?). You can check Google Trends to
get some insights:

https://trends.google.es/trends/explore?date=2015-01-01%202017-12-31&q=lisboa,algarve&hl=es

```{r max_dia}
max(table(x$dia))
which.max(table(x$dia))
```
With the computed day 'dia', along with the variables 'stays_in_week' and 
'weekend_nights', we can try to manually categorize the trip type according
to the following criteria (this is arbitrary, clearly improvable):

1) if 'stays_in_weekend_nights' is zero => work trip
2) if 'stays_in_week_nights' is zero or one and in this case the entry is on
   Friday => weekend
3) if 'stays_in_week_nights' is five and 'stays_in_weekend_nights' is three 
  (that is, from Saturday or Sunday to Saturday or Sunday) 
   => week holiday package
4) if 'stays_in_weekend_nights' is one or two and 'stays_in_week_days' is five
   or less => work + rest
5) the rest of combinations => holidays

```{r tipo_visita}
# require(lubridate)
x$tipo=ifelse(x$stays_in_weekend_nights==0, "work",
       ifelse(x$stays_in_week_nights==0, "weekend",
       ifelse(x$stays_in_week_nights==1 & wday(x$dia)==6, "weekend",
       ifelse(x$stays_in_week_nights==5 & 
              (x$stays_in_weekend_nights==3 |
               x$stays_in_weekend_nights==4), "package",
       ifelse(x$stays_in_week_nights<=5 & 
              x$stays_in_weekend_nights<3, "work+rest",
       "rest")))))
```

One way to refine this classification would be to look at the number of adults,
children, and infants to decide whether it is a business traveler or a family.
The possibilities are endless: you can enrich the dataset with geographic data
(distance between countries), demographic data, economic data (per capita
income), weather data (in both Portugal and the country of origin), etc. 

EXERCISE: You must explore such enriched dataset and, in this process of 
exploration, decide what story you want to tell about it. Some ideas:

1) do tourists from different countries travel in different dates?
2) differences in cancellations among groups (countries, type of stay, ...)
3) relationship between type of stay 'tipo' and cost 'adr'
4) differences among groups with respect to hotel type (city / resort)

```{r final_storytelling_exercise}
# STORY: Relationship between Stay Type ('tipo') and Booking Cost ('adr')
# We want to see if "Package" deals are actually cheaper per day than "Work" trips.

ggplot(x, aes(x = reorder(tipo, adr, FUN = median), y = adr, fill = tipo)) +
  geom_boxplot(outlier.shape = NA) +
  coord_cartesian(ylim = c(0, 300)) +
  labs(title = "Economic Analysis of Stay Types",
       subtitle = "Median ADR across different trip categories",
       x = "Stay Type", y = "ADR (Euro)") +
  theme_minimal() +
  theme(legend.position = "none")

# Statistical confirmation
ggbetweenstats(data = x, x = tipo, y = adr, 
               title = "Statistical comparison of costs per stay type")

# Findings: Work trips and short "Work+Rest" stays tend to have higher daily rates 
# than long-term packages, suggesting business travel is a higher-yield segment.
```

NOTE: This is a good example of using ChatGPT or other generative AI to ask 
interesting questions about the proposed dataset. The following paper describes
the potential uses of generative AI in the different phases of creating a data
visualization for storytelling:

https://ieeexplore.ieee.org/stamp/stamp.jsp?arnumber=10891192
